Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-07-24 Manuscript/manuscript.Rmd
manuscript edits

Genomic annotation data in the form of cell type specific chromatin states, histone marks, and transcription factor binding sites (GRCh37/hg19 human genome assembly) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 2). Functional annotations were represented by chromosome-specific regions on the linear genome obtained by ChIP-seq peak calling algorithms.

@@

A total of 78 cell type specific functional genomic annotations derived from ChIP-Seq experiments (See Methods) for the GM12878 and K562 cell line were used to build the feature space of downstream predictive models. We constructed three different predictor types for each functional genomic annotation. Counts of overlaps (**overlap count**, OC) were defined as the total number of peak regions that overlapped within a given genomic bin (Figure 3A). Percents of overlaps (**overlap percent**, OP) were defined as the proportion of a bin overlapped by a peak region (Figure 3B). **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations (Figure 3C). A log2-transformation of distance was used to account for the skewness present in the distributions of distances (Supplementary Figures 1-2). Each predictor type provides a granular way to draw associations between functional genomic annotations and the spatial structure of the genome.

@@

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models [??? Never defined, what is this?; !!!defined above; should i mention again?], within each cell line and across resolution. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites). We found that for cell line specific random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the functional genomic annotations classified as transcription factor binding sites (TFBS) (Figure 6 A). This was consistent across resolutions, suggesting that TFBS were the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

@@

>>>>

Genomic annotation data in the form of cell type specific chromatin states, histone marks, and transcription factor binding sites (GRCh37/hg19 human genome assembly) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Tables 2-3). Functional annotations were represented by chromosome-specific regions on the linear genome obtained by ChIP-seq peak calling algorithms.

@@

A total of 78 cell type specific functional genomic annotations derived from ChIP-Seq experiments (See Methods) for the GM12878 and K562 cell line were used to build the feature space of downstream predictive models (Supplementary Tables 2-3). We constructed three different predictor types for each functional genomic annotation. Counts of overlaps (**overlap count**, OC) were defined as the total number of peak regions that overlapped within a given genomic bin (Figure 3A). Percents of overlaps (**overlap percent**, OP) were defined as the proportion of a bin overlapped by a peak region (Figure 3B). **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations (Figure 3C). A log2-transformation of distance was used to account for the skewness present in the distributions of distances <!--(Supplementary Figures 1-2)-->. Each predictor type provides a granular way to draw associations between functional genomic annotations and the spatial structure of the genome.

@@

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models, within each cell line and across resolution. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites). We found that for cell line specific random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the functional genomic annotations classified as transcription factor binding sites (TFBS) (Figure 6 A). This was consistent across resolutions, suggesting that TFBS were the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

@@

## Top TFBS were more enriched at predicted TAD boundaries at bp resolution than those provided by the TAD caller kb resolution

We then shifted our focus toward developing a method to more accurately identify the location of TAD boundaries throughout the linear genome. Typical TAD callers are restricted to kilobase resolutions, which increase the variability in the actual location of a boundary. Instead, we aimed to predict the location of TAD boundaries at basepair resolution using optimally built cell line specific random forest classifiers. Each model was restricted to only include the most influential cell line specific TFBS, namely CTCF, RAD21, SMC3a, adn ZNF143. We then validated our models on cell line and chromosome specific regions of the genome, at base pair resolution. Predicted TAD boundary regions (PTBR) were identified as the collection of base pairs where the predicted probability vector was equal to 1, which denoted 100% accuracy in correctly identifying a basepair as a potential TAD boundary (Figure 7A). To allow for minor fluctuations in probability, any base pair given probability of 1 that was within a 10 bp window from another base pair also given probability of 1 were combined into one PTBR. PTBR were then compared to the called TAD boundary regions (CTBR), which here were defined by a 10 kb resolution flanking (See Methods). There were a total of 4663 identified PTBR compared to 237 CTBR. We found that approximately 59% (139 out of 237) of CTBR were overlapped by PTBR, indicating there are fluctuations in probabilities within the 20 kb regions defining CTBR (Figure 7B). We then compared distances between PTBRs and annotations versus CRBRs and annotations. We found that boundary regions predicted by our model at bp resolution experienced short distance profiles to known architectural proteins than did boundary regions called by Arrowhead.

![**Comparing TAD boundaries regions predicted at bp resolution with those called at 10 kb resolution.** (A) A subset of the probabilitity vector for predicting TAD boundaries for chr22:17380000-chr22:18650000. The red dashed lines represent predicted TAD boundary regions, while the blue dashed lines represent called TAD boundaries identified by Arrowead at 10 kb. A called TAD boundary region is one flanked by 10 kb (B) Venn diagram illustrating the number of overlapped regions between predicted (red) and called (blue) TAD boundary regions. (C) Boxplots illustrating the distance profiles between each class of boundary regions (predicted-red; called-blue) and architechtural proteins known to be enriched at TAD boundaries.](Figures/pred_at_bp_res_plots.png)
